---
title: "Writing a configuration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Writing a configuration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE, error = FALSE)
```

One YAML file describes a whole run, and `platypus_spec()` reads it. This is the walk
through that file; the exhaustive list of fields lives on the engine's site, because the
file is the engine's format and keeping a second copy here would mean two of them going
out of step:
[every configuration field](https://maju116.github.io/pyplatypus/config.html).

**The same file works in both packages.** R reads it with `platypus_spec()`, Python with
`from_yaml()`, and both hand it to the same validator. Nothing below is R-specific except
the lines that run it.

Arguments and YAML produce the same object and nothing downstream can tell which was used,
so a configuration file and an R script describe the same experiment - and either can be
handed to a colleague.

## One file, one run

The shortest thing that trains:

```yaml
task: semantic_segmentation
data:
  train_path: data/train
  validation_path: data/validation
  colormap: [[0, 0, 0], [255, 255, 255]]
models:
  - name: first_try
    input_shape: [256, 256]
```

```{r}
spec <- platypus_spec("run.yaml")
fit <- platypus_fit(spec)
evaluate(fit)
```

Four things are required and the rest have defaults: `task`, `data.train_path`, and a
`name` and `input_shape` for each model. Everything else - the architecture, the loss, the
optimiser, how many epochs - has a default that lives in the engine, so **a short
configuration is a complete one** rather than an incomplete one. `?u_net` has a *Where the
defaults live* section saying why they are not written into the R signatures.

`colormap` says which colour in a mask means which class, darkest first. Masks stored as
label maps rather than pictures use `labels: [0, 1]` instead, and giving both is an error:
exactly one says how a mask is encoded.

## Several models at once

`models:` is a list, and that is the reason this file format exists rather than only the
code path. Each entry trains on the same data and lands in the same table, so comparing
variants is one run rather than a loop somebody has to write.

```yaml
task: semantic_segmentation
data:
  train_path: data/train
  validation_path: data/validation
  colormap: [[0, 0, 0], [255, 255, 255]]
models:
  - name: shallow
    input_shape: [256, 256]
    blocks: 3
    filters: 16
  - name: deep
    input_shape: [256, 256]
    blocks: 5
    filters: 32
    architecture: u_net_plus_plus
```

```{r}
fit <- platypus_fit(platypus_spec("run.yaml"))
evaluate(fit)            # one row per model
training_history(fit)    # one row per model per epoch
```

`name` has to be unique, because it is how a model is identified in every score table, in
every record and in `predict(fit, model = ...)`. Two models called the same thing is
refused rather than resolved, by name:
`model names must be unique; repeated: shallow`.

Two entries differing only in `blocks` is a legitimate configuration and a common one: this
is for comparing variants, not only architectures.

## Dividing one folder

With no `validation_path`, say how to divide `train_path`:

```yaml
task: semantic_segmentation
data:
  train_path: data/all
  split:
    fractions: [0.7, 0.15, 0.15]
    group_by: "^(patient\\d+)_"
  colormap: [[0, 0, 0], [255, 255, 255]]
models:
  - name: by_patient
    input_shape: [256, 256]
```

Two fractions give train and validation, three add a test set. Nothing is written to disk -
the samples are divided after they are discovered, so there are no new folders and no CSV.
`split_dataset()` is still the tool when the CSVs themselves are the point.

**`group_by` is required and may be `NULL`.** It is a regular expression whose first group
is the thing that must not appear on both sides - a patient, a scan, a slide. The mistake it
prevents leaves no trace: slices of one patient in both sets make validation measure memory
rather than generalisation, and every number in the output still looks reasonable. An
omitted key would be that choice made silently; `null` is the same choice made on purpose,
for data where every sample is independent.

Exactly one of `split` and `validation_path`. If a run genuinely has no validation set - a
final fit on everything - say `validation: false`, which is the third answer rather than an
absence, because "I have no validation set" and "I forgot" look identical otherwise.

The same from R, where the argument answers the same question three ways:

```{r}
segmentation_data("data/all", split = list(fractions = c(0.8, 0.2), group_by = NULL))
segmentation_data("data/train", validation = "data/validation")
segmentation_data("data/all", validation = FALSE)
```

## Augmentation, callbacks, and the rest of a model

```yaml
task: semantic_segmentation
data:
  train_path: data/train
  validation_path: data/validation
  colormap: [[0, 0, 0], [255, 255, 255]]
models:
  - name: augmented
    input_shape: [256, 256]
    epochs: 40
    loss: {name: dice}
    metrics: [{name: iou}, {name: dice}]
    optimizer: {name: adam, learning_rate: 0.001}
    callbacks:
      - {name: early_stopping, monitor: val_loss, patience: 10}
      - {name: model_checkpoint, monitor: val_iou, path: runs/best.safetensors}
    augmentation:
      - {name: HorizontalFlip, params: {p: 0.5}}
      - {name: RandomBrightnessContrast, params: {p: 0.3}}
```

Every one of those blocks has an R constructor that builds the same thing, and the help
pages for them are where the measurements are: `?losses`, `?metrics`, `?optimizers`,
`?callbacks`.

```{r}
u_net(
  "augmented",
  input_shape = c(256, 256),
  epochs = 40,
  loss = loss_dice(),
  metrics = list(metric_iou(), metric_dice()),
  optimizer = optimizer_adam(learning_rate = 0.001),
  callbacks = list(
    callback_early_stopping(monitor = "val_loss", patience = 10),
    callback_model_checkpoint(monitor = "val_iou", path = "runs/best.safetensors")
  ),
  augmentation = list(
    augmentation_step("HorizontalFlip", p = 0.5),
    augmentation_step("RandomBrightnessContrast", p = 0.3)
  )
)
```

**There is no `mode` on a callback that watches a number.** The direction is taken from the
name: anything ending in `loss` is better lower, anything else better higher. So
`monitor = "val_iou"` maximises and `monitor = "val_loss"` minimises, and there is nothing
to set inconsistently.

**Augmentation applies to the training data only** - never to validation, where the point is
to measure the same thing every epoch. The names are albumentations' own, and
`available_transforms()` lists what the installed version offers; at `rank = 3` the list is
shorter, because support for volumes is uneven. Each step is probed at the model's own input
size while the pipeline is built, so a transform that cannot do the job says so before the
first epoch rather than during it.

## Detection instead of segmentation

`task: object_detection` changes which fields the file accepts. Boxes rather than masks, so
`classes` and an annotation format rather than a colormap:

```yaml
task: object_detection
data:
  train_path: data/train
  validation_path: data/validation
  annotation_format: pascal_voc
  classes: [rbc, wbc, platelets]
models:
  - name: cells
    input_shape: [416, 416]
```

**Leave `anchors` out and they are fitted to your own boxes** by k-means on an IoU distance,
`anchors_per_grid` of them per grid. That is worth about 0.23 of mean overlap against COCO's
nine on blood cells - measured, 0.65 against 0.88 - so the default is the useful one and
naming anchors by hand is for reproducing a published model. `detection_anchors()` reports
them and `plot_anchors()` draws them against the boxes they were fitted to.

## Not training at all

```yaml
task: semantic_segmentation
data:
  train_path: data/images
  validation_path: data/images
  colormap: [[0, 0, 0], [255, 255, 255]]
models:
  - name: published
    input_shape: [256, 256]
    weights: dsbowl-unet
    fit: false
```

`weights` takes a registry name, a local path, or `hf://owner/repo/file@commit` - and the
`hf://` form requires the commit, because weights that change under a stable name are the
worst kind of irreproducibility. A registry name carries one already;
`available_weights()` lists them.

With `fit: false` nothing trains and `predict(fit)` works immediately. The weights are
checked against the model they are being loaded into *before* torch is touched: a shape, a
channel count or a class count that does not match is a refusal naming the field, not a
tensor-size error from inside a loss.

## What is checked, and when

Validation happens in this order, and the first two cost no epochs:

1. **The file, against the schema.** Unknown fields, missing required ones, a `name:` that no
   block has, a callback watching a number nothing will produce. Before any path is read -
   and before Python starts, for the checks R can make itself.
2. **The data on disk**, when the run begins: that the folders exist, that images and masks
   correspond, and that every class in `colormap` or `labels` actually appears in the masks,
   because a declared class appearing in none has a target channel of zeros, so the model is
   never shown what it is being asked to find. `inspect_masks()` runs that check on its own,
   and `platypus_fit(check_masks = FALSE)` skips it.
3. **Everything else**, during the run.

A refusal names the field and says what to do. The engine's own messages are translated into
R conditions rather than passed through as Python tracebacks, so
`tryCatch(platypus_spec("run.yaml"), error = function(e) conditionMessage(e))` gives a
sentence rather than a stack.

## The same run from code

The file is not a second implementation. These describe the same experiment:

```{r}
spec <- platypus_spec("run.yaml")

spec <- platypus_spec(
  data = segmentation_data(
    "data/train",
    validation = "data/validation",
    colormap = list(c(0, 0, 0), c(255, 255, 255))
  ),
  models = list(u_net("first_try", input_shape = c(256, 256)))
)
```

And `as.list(spec)` is the file, so a run built from arguments can be written out and kept:

```{r}
yaml::write_yaml(as.list(spec), "run.yaml")
```

Which to use is a question about how many models and how often, not about what is possible.
A file is better when there are several models, when the run is going to be repeated, or
when somebody else has to see what was done; the arguments are better inside a loop over
something the file cannot express, like a vector of datasets.

Note that `as.list()` is also how the fields are reached: the object returned by
`platypus_spec()` carries the specification on the Python side, so `spec$task` is `NULL`
while `as.list(spec)$task` is what you want.
